Dynamic Walking and Whole-Body Motion Planning for Humanoid Robots: an Integrated Approach

### 2012-11-27 paper.tex
Fix some syntax and grammar errors.

\author{
  S\'ebastien Dalibard\and
  Antonio El Khoury\and
  Florent Lamiraux\and
  Alireza Nakhaei\and
  Michel Ta\"ix\and
  Jean-Paul Laumond
  \footnote{The authors are with CNRS ; LAAS ; 7 avenue du colonel Roche, F-31077
    Toulouse Cedex 4, France and Universit\'e de Toulouse ; UPS, INSA, INP, ISAE ;
    UT1, UTM, LAAS ; F-31077 Toulouse
    Cedex 4, France. S\'ebastien Dalibard is now with Nanyang Technological University, Singapore.}
  \footnote{
    This paper summarizes and and extends previous work that appeared in the 9th and 11th
    IEEE-RAS International Conference on Humanoid Robots, 2009 and 2011.
  }
}

@@

This paper presents a general method for planning collision-free whole-body walking motions
for humanoid robots. First, we present a randomized algorithm for constrained motion
planning, that is used to generate collision-free statically stable paths solving
manipulation tasks. Then, we show that dynamic walking makes humanoid robots 
small-space controllable. Such a property allows to easily transform 
collision-free statically stable paths into 
collision-free dynamically balanced trajectories. It leads to a sound 
algorithm which has been applied and evaluated on several
problems where whole-body planning and  walk are needed,
and the results have been validated  on a real HRP-2 robot.

@@

This work is based on several fields of humanoid robotics
research: prioritized inverse kinematics, randomized  whole-body  motion
planning and walk pattern generators based on the Zero-Momentum
Point (ZMP) formalism. This section summarizes the literature related to each of 
these fields.

@@

When  planning a  whole-body motion  for a  humanoid robot, one difficult
challenge is to cope with  the curse of dimensionality. The complexity
of   motion  planning  is   exponential  in   the  dimension   of  the
configuration  space ($\mathcal{C}$)  to explore.  When  dealing with
high-dimensional configuration  spaces, it is  typically impossible to
explicitly represent  them, leading to the use  of randomized sampling
techniques  to solve  global planning  problems. In  the  past fifteen
years,  \textit{Probabilistic Roadmaps} \cite{kavraki1996prp} and  
\textit{Rapidly exploring Random  Trees} (RRT) 
\cite{kuffner00rrtconnect}  have been  developed and  used to  solve many
high-dimensional   planning  problems, see \cite{Lav06} and \cite{choset2005prm} for comprehensive
overviews.
When  using  sampling  techniques  on  a humanoid  robot,  another difficulty
is to  take into  account stability  constraints,  i.e. to
generate  random  configurations   on  zero  volume  submanifolds  of
$\mathcal{C}$. This problem has been investigated with success during
the last few years, \cite{Berenson15032011} presents a exhaustive survey
of Jacobian-based methods. Other recent contributions \cite{porta2012randomized}
present sophisticated constrained planning techniques based on higher-dimensional
continuation. Section \ref{sec:wb} presents a simple adaptation
of the RRT algorithm to constrained motion planning, that was first
introduced in \cite{dalibard09}.

@@

Another  field of  humanoid  robotics research  is  the generation  of
dynamically stable  walk patterns. Since  the introduction of  the ZMP
formalism  \cite{vukobratovic1969contribution},  several  methods  have  been  proposed  to
generate  walking  motions efficiently.   One  way  to  deal with  the
complexity  of a  humanoid  robotics  kinematic tree  is  to use  the
so-called "cart-table" simplified model \cite{kajita2003biped}. Based on this
model,  planning a trajectory  for the  ZMP is  reduced to  planning a
trajectory  for  the Center  of  Mass (CoM)  of  the  robot.  Given  a
trajectory  of  the CoM  and  footstep  positions, inverse  kinematics
solvers can animate  the whole set of degrees of freedom (DoFs) of the
robot to generate a dynamically stable walk trajectory.

@@

Collision-free  locomotion   trajectories  are  usually   obtained  by
simplifying the model of the  robot or of the environment. By reducing 
a  humanoid robot to a bounding volume that wraps the swaying motion,
one can  use a simple  planar motion planner  on this bounding  volume and
generate  a valid  locomotion  trajectory. This  strategy  is used  in
\cite{pettre20032} in a computer animation context. Variants of this method
include  dynamic path reshaping  \cite{yoshida-humanoids05}: if  collisions appear
when animating  the locomotion  trajectory, it is  locally reshaped
and re-animated.  This two-stage  strategy does not guarantee that the
locomotion trajectory can be followed or that the local reshaping will
converge.

@@

Some planning methods for free-climbing robots \cite{bretl2006motion}
can be seen as a general way 
to consider quasi-static multi-step planning. They are not directly applicable to
humanoid dynamically stable locomotion.
Other recent contributions to the field of locomotion planning  include algorithms 
considering the dynamics at the planning phase \cite{shkolnik2011bounding}. This leads to 
a growth of algorithmic complexity, particularly costly for high-dimensional
systems such as humanoid robots. To the authors knowledge such techniques have not been
used on humanoid robotic platforms so far.
In our work, we show that it is not always
necessary to take into account dynamics while planning, even though the final
output of our algorithm is a dynamically stable walk motion. This result was first
presented in \cite{dalibard2011small}.

@@

The main contribution of this work is a  whole-body motion planner for humanoid robots
that computes collision-free walking trajectories, based on
exact models of both robot and environment. It is used to solve manipulation tasks
that may require walking. The first stage of our algorithm uses a
constrained sampling-based motion 
planner and  computes a collision-free  statically stable path  for a robot fixed or sliding 
on the ground. Another  contribution of this  paper is the formal  proof that
this first path can always be  approximated by a dynamically balanced, collision-free
walking  trajectory. We have implemented this well-grounded 
method.  The results have been validated on a real HRP-2 robot.

\section{Randomized Motion Planning on Constrained Manifolds}
\label{sec:wb}

This section presents an algorithm for constrained motion planning. 
The problem solved here
differs from classic approaches in two ways:
\begin{enumerate}
\item the set of valid configurations is defined implicitly, as the ones satisfying a given 
  set of inverse kinematics and balance tasks;
\item the goal is also defined implicitly, by additional inverse kinematic tasks.
\end{enumerate}
During global planning, we will consider several types of
tasks and constraints for various reasons:
\begin{itemize}
\item Static stability: the center of mass of the robot stays above
  the support polygon center, the two feet have fixed position and
  orientation.
\item End-effector position and orientation: the goals of some problems presented
  in the experimental section of this paper are defined as robot hand configuration,
  or gaze direction.
\item Configuration task: our adaptation of randomized motion planning algorithms 
  uses tasks defined as the distance towards a given configuration in $\mathcal{C}$. 
  This will be detailed in the following section.
\end{itemize}

@@

The classic RRT algorithm, as presented in  \cite{kuffner00rrtconnect}, grows 
a random tree inside the robot 
collision-free configuration space 
$\mathcal{C}_{free}$. Each iteration of the algorithm attempts to extend the tree
by adding new vertices in the direction of a randomly-selected configuration
$q_{rand}$. Algorithm~\ref{algRRT} shows the pseudo-code of the RRT algorithm.
It takes as input an initial configuration $q_0$ and grows a tree  $\mathcal{T}$ rooted 
in $q_0$.

@@

Section~\ref{sec:goal-sampling}  describes the goal manifold sampling, and 
Section~\ref{sec:extension} the adaptation or RRTs random extensions to constrained planning.

@@

The way we generate a goal configuration is the following:
\begin{enumerate}
\item Shoot a random configuration in $\mathcal{C}$ with uniform distribution.
\item Call the \textit{constraint solver} (Algorithm~\ref{algo:newton}) on this configuration, with the static
  stability constraints and the goal task.
\item If success, check for collisions.
\end{enumerate}

@@

Fig.~\ref{fig:goal} shows different random configurations respecting stability 
constraints and solving a reaching task for an animated character.

@@

\caption{Random goal configurations solving a reaching task. All the
  configurations are stable and collision-free, and the right hand of the character
  reaches the orange ball. Note that the ball is also an obstacle.}
\label{fig:goal}
\end{figure}

@@

We present in Fig.~\ref{fig:wb-shelves} an illustration of 
the use of randomized motion planning 
on complex manipulation problems. The humanoid robot HRP-2 faces shelves. It has to: (i) grasp
a ball lying on a shelf, (ii) put it on a higher shelf, (iii) come back to a natural
rest configuration. The statically stable manifold on which the planning takes place is
defined by tasks constraining the robot feet and CoM. The goal of the first problem
is defined by a hand configuration task (3 positions and 2 orientations), and a gaze task
(the robot has to look at the ball). The goal of the second problem is defined by a 
different hand configuration and the gaze task.
Finally, the goal of the third problem is the rest configuration.

@@

\caption{HRP-2 grabs a ball on a shelf, puts it on another shelf, and comes back to 
  a rest position. Static stability constraints are respected along the path, and 
  the intermediary goals consisting in grasping and displacing the ball are defined
  implicitly as inverse kinematic tasks.}
\label{fig:wb-shelves}
\end{figure}

We have run this set of motion planning problems 20 times. The results are compiled in Table~\ref{table:reaching}.
\begin{table}
\begin{tabular}{lrrrr}
& min & max & average & average \\ 
&&&& per problem \\
number of nodes & 43.00 & 481.00 & 102.70 \\
goal generation time \hspace {1.cm} & 1.00 & 1.56 & 1.22\\
planning time& 67.36 & 376.84 & 134.28 & 44.76\\
\end{tabular}
\caption {Experimental results on 20 runs: 3 motion planning problems and 2 goal generations. Times are expressed in seconds.}
\label{table:reaching}
\end{table}

@@

The previous section has presented a simple algorithm that solves manipulation planning
problems on a given constrained manifold of $\mathcal{C}$.

If we use this algorithm with static 
stability constraints without fixing globally the robot foot positions, it generates 
statically stable paths for a robot sliding on the ground. Fig~\ref{fig:sliding} shows
an example of a whole-body collision-free path for a robot passing between two chairs.
Such paths are physically unfeasible. They are, however, easier to generate than feasible
dynamic trajectories because only geometric constraints are considered at planning time.

@@

  \caption{Collision-free  statically stable path  for a  humanoid  robot  sliding on  the
    ground.}
  \label{fig:sliding}
\end{figure}

This section presents a \textit{constructive} proof that any such statically stable, 
collision-free path for a legged robot sliding
on the ground can be approximated by a collision-free dynamic walk trajectory.
The proof is based on ideas from control theory, in particular small-space 
controllability. It also uses the fact that balance criteria for dynamic walk are different
from the ones for static stability. An important point is that our result and method do not hold
for a biped robot that walks in a quasi-static way.

Section~\ref{sec:ssc} recalls the definition of small-space controllability and its
use in motion planning. Section~\ref{sec:humanoid-ssc} proves that a legged robot
dynamically walking is small-space controllable, while a legged robot quasi-statically
walking is not. Section~\ref{sec:ssc-application} shows how this property is used to
approximate collision-free statically stable paths by dynamic walk trajectories.

@@

  \caption{The small-space controllability local property:  any configuration $q'$ 
    at distance less than
    $\eta$ is reachable from $q$ by an admissible trajectory included in
    a ball of size $\epsilon$.}
  \label{fig:ssc1}
\end{figure}

The main  consequence of  that property  
in  motion planning  is the following theorem, that shows how planning for
dynamic systems is reduced to geometric planning
thanks to small-space controllability:

@@

Fig. \ref{fig:ssc2} shows an example of collision-free 
path approximation by admissible collision-free sub-trajectories. The fact 
that this algorithm  converges is guaranteed by the small-space
controllability property.

@@

\begin{proof}[Proof of Theorem \ref{thm:ssc}]
  Let $\mathcal{C}$ be the configuration space of a small-space controllable robot, and 
  $\mathcal{C}_{free} \subset \mathcal{C}$ the set of collision-free configurations. We
  consider in-contact configurations as colliding, so $\mathcal{C}_{free}$ is an open set.
  Let $\tau : [0,1] \rightarrow \mathcal{C}_{free}$ be a collision-free path. For all $x \in [0,1]$,
  $\tau(x) \in \mathcal{C}_{free}$, so there exists $\epsilon_x$ such that the open ball 
  $B(\tau(x),\epsilon_x)$ of center $\tau(x)$ and radius $\epsilon_x$ is included in 
  $\mathcal{C}_{free}$. The small-space controllability property states that for all $x$,
  there exists $\eta_x > 0$ such that every configuration $q \in B(\tau(x),\eta_x)$ is reachable 
  from $\tau(x)$ by a trajectory included in $B(\tau(x),\epsilon_x)$.

@@

This section  discusses walking robot small-space controllability. To clarify
the presentation, we consider a simplified model of a legged robot consisting of two feet
of zero mass and a point mass free to move in three dimensions.
We do not consider 
the kinematic chains between the feet and the mass. The robot is walking on a flat terrain,
and the feet are assumed to have a positive surface. For our presentation, it is not 
necessary to consider the feet height, so the configuration space of the robot is:
\[
\mathcal{C} = SE(2) \times SE(2) \times \mathbb{R}^3
\]
It is of dimension 9.

@@

\begin{theorem}
\label{thm:humanoid-ssc}
A quasi-static walking robot is not small-space controllable. A dynamic walking robot is.
\end{theorem}

@@

The first claim is straightforward. Let the robot be in a configuration $q$ where the two
feet are separated by a positive distance. Let  $L>0$ be the positive horizontal distance
between the CoM and the left foot (if the CoM is over the left foot, we can consider similarly 
the right foot). For all $\epsilon < L$, any valid trajectory starting from $q$, included in
the ball of center $q$ and radius $\epsilon$, is such that the CoM is never over the left foot.
Given the quasi-static walk conditions, the right foot of the robot is fixed along 
the trajectory. Thus, the set of accessible configurations from $q$ by staying inside 
$B(q,\epsilon)$ does not form a neighborhood of $q$, since it does not include any configuration
where the right foot has moved. This shows that the robot is not small-space controllable.

@@

Let us now consider a dynamic walking robot. Starting from any valid static configuration, 
the CoM can move vertically without affecting balance, so there is no need to 
consider this degree of freedom in the following. If the CoM is not over the edge of the support
polygon, it is possible to move it in a quasi-static way inside a neighborhood of its current position that projects itself over the support polygon.
It is thus sufficient and necessary to prove that for all $\epsilon >0$, it is possible to move
the feet while keeping the CoM inside a neighborhood of size $\epsilon$. Let such $\epsilon >0$
be arbitrarily fixed.

@@

Without loss of generality, let us assume that the robot is in a configuration
in which the CoM is at the horizontal position $(0,0)$, the foot centers  are 
aligned with the $y$-axis and the horizontal distance between the CoM and either of the foot centers
is $L$. To achieve dynamically balanced walking, we aim at making
$p_y(t)$ oscillate  between $-L$ and $L$. To move the ZMP  under a
given foot, only  the $y$ coordinate of the CoM  is of interest. Thus,
we will keep the $x$ coordinates  of the CoM and ZMP constant equal to
$0$. By hypothesis, the feet are of positive surface,
let $l>0$ be such that the length of the section of a foot along the 
$y$-axis is greater than $l$. Fig.~\ref{fig:simple-humanoid} summarizes the notations used
in the following.

@@

Starting from a static configuration at time $(t=0)$, we cannot apply
directly  a  command  $y(t)  =  \epsilon \sin(\omega  t)$  because  it
generates  a discontinuity  in the  speed of  the CoM  at time $(t=0)$. To
overcome this  discontinuity, we go through a  transient state between
$(t=0)$ and  $(t=T)$ for some  $T >0$. Let  $f:[0,T] \rightarrow
[0,1]$  be an  increasing function of class $C^\infty$  such  that  $f(0)  =  0$,
$\dot{f}(0) = 0$, $f(T) =  1$, $\dot{f}(T) = 0$ and $\ddot{f}(T)
=  0$.  We can explicitly  construct  such  an $f$   with  a degree  4
spline.   We   also   request   that   for  all   $t   \in   [0,T]$,
$|2\epsilon\dot{f}(t)\frac{\omega}{\omega_0^2}|   \leq  \frac{l}{4}$
and   $|\epsilon\ddot{f}(t)/\omega_0^2|  \leq   \frac{l}{4}$.  These
inequalities  will be  used  to bound  the  trajectory of  the  ZMP. We  can
guarantee them by  choosing $T$ large enough. Let  us now consider the
following CoM motion:

@@

For    all     $t    \in    [0,T]$,    $f(t)     \epsilon    (1    +
\frac{\omega}{\omega_0}^2)  \sin  (\omega t)$  lies  between $-L$  and
$L$. The bounds on the derivatives of $f$ guarantee that $p_y(t)$ lies
between  $-L- l/2$ and  $L+ l/2$,  which means  that the  ZMP stays
inside  the  support  polygon.  Fig.  \ref{fig:zmp-inplace}  shows  an
example  of CoM  motion  on the  $y$  axis and  the corresponding  ZMP
motion. Once in permanent walk in-place state, the robot can come back
to  a static  state by  applying a  symmetric transient  state  used to
decrease  gradually  the amplitude  of  the  oscillations  of the  CoM
without  generating a  discontinuity in  the first  derivative  of the
command.

@@

\subsubsection{Remarks}
\paragraph{Generalization to a complete model:} More work is needed to generalize 
the previous proof to any legged robot model. First, because we use a model with
a point mass, we did not have to consider the angular-momentum of the robot
around the CoM, while it accounts for dynamic balance. Moreover, when considering
an actual kinematic and dynamic model,
the positions of feet and CoM are not controlled degrees of freedom, but inputs 
for inverse kinematic tasks. Continuity and regularity results about the robot
controller -- i.e. how much the configuration changes when the input of an inverse
kinematics task changes--, would be needed to generalize the proof. We have not
proven such properties for the inverse kinematics controller we use. 
The  proof presented above still gives the intuition for the
algorithm presented in the following.

@@

In the problem shown in Fig. \ref{fig:shelf} the robot has to put a ball on a
shelf, in a constrained apartment environment. The final configuration is defined 
implicitly as a desired hand position. We have generated automatically goal configurations 
solving the task, as described in Section~\ref{sec:wb} . Then, we have
applied our planner to generate a whole-body walk motion that solves the hand reaching
task.

>>>>

\author{
  S\'ebastien Dalibard\and
  Antonio El Khoury\and
  Florent Lamiraux\and
  Alireza Nakhaei\and
  Michel Ta\"ix\and
  Jean-Paul Laumond
  \footnote{The authors are with CNRS ; LAAS ; 7 avenue du colonel Roche, F-31077
    Toulouse Cedex 4, France and Universit\'e de Toulouse ; UPS, INSA, INP, ISAE ;
    UT1, UTM, LAAS ; F-31077 Toulouse
    Cedex 4, France. S\'ebastien Dalibard is now with Nanyang Technological University, Singapore.}
  \footnote{
    This paper summarizes and extends previous work that appeared in the 9th and 11th
    IEEE-RAS International Conference on Humanoid Robots, 2009 and 2011.
  }
}

@@

This paper presents a general method for planning collision-free whole-body walking motions
for humanoid robots. First, we present a randomized algorithm for constrained motion
planning, that is used to generate collision-free statically stable paths solving
manipulation tasks. Then, we show that dynamic walking makes humanoid robots 
small-space controllable. Such a property allows to easily transform 
collision-free statically balanced paths into 
collision-free dynamically balanced trajectories. It leads to a sound 
algorithm which has been applied and evaluated on several
problems where whole-body planning and  walk are needed,
and the results have been validated  on a real HRP-2 robot.

@@

This work is based on several fields of humanoid robotics
research: prioritized inverse kinematics, randomized  whole-body  motion
planning and walk pattern generators based on the Zero-Moment
Point (ZMP) formalism. This section summarizes the literature related to each of 
these fields.

@@

When  planning a  whole-body motion  for a  humanoid robot, one difficult
challenge is to cope with  the curse of dimensionality. The complexity
of   motion  planning  is   exponential  in   the  dimension   of  the
configuration  space ($\mathcal{C}$)  to explore.  When  dealing with
high-dimensional configuration  spaces, it is  typically impossible to
explicitly represent  them, leading to the use  of randomized sampling
techniques  to solve  global planning  problems. In  the  past fifteen
years,  \textit{Probabilistic Roadmaps} \cite{kavraki1996prp} and  
\textit{Rapidly exploring Random  Trees} (RRT) 
\cite{kuffner00rrtconnect}  have been  developed and  used to  solve many
high-dimensional   planning  problems, see \cite{Lav06} and \cite{choset2005prm} for comprehensive
overviews.
When  using  sampling  techniques  on  a humanoid  robot,  another difficulty
is to  take into  account balance  constraints,  i.e. to
generate  random  configurations   on  zero  volume  submanifolds  of
$\mathcal{C}$. This problem has been investigated with success during
the last few years, \cite{Berenson15032011} presents an exhaustive survey
of Jacobian-based methods. Other recent contributions \cite{porta2012randomized}
present sophisticated constrained motion planning techniques based on higher-dimensional
continuation. Section \ref{sec:wb} presents a simple adaptation
of the RRT algorithm to constrained motion planning, that was first
introduced in \cite{dalibard09}.

@@

Another  field of  humanoid  robotics research  is  the generation  of
dynamically balanced walk patterns. Since  the introduction of  the ZMP
formalism  \cite{vukobratovic1969contribution},  several  methods  have  been  proposed  to
generate  walking  motions efficiently.   One  way  to  deal with  the
complexity  of a  humanoid  robot  kinematic tree  is  to use  the
so-called "cart-table" simplified model \cite{kajita2003biped}. Based on this
model,  planning a trajectory  for the  ZMP is  reduced to  planning a
trajectory  for  the Center  of  Mass (CoM)  of  the  robot.  Given  a
trajectory  of  the CoM  and  footprint  positions, inverse  kinematics
solvers can animate  the whole set of DoFs of the
robot to generate a dynamically stable walk trajectory.

@@

Collision-free  locomotion   trajectories  are  usually   obtained  by
simplifying the model of either the  robot or the environment. By reducing 
a  humanoid robot to a bounding volume that wraps the swaying motion,
one can  use a simple  planar motion planner  on this bounding  volume and
generate  a valid  locomotion  trajectory. This  strategy  is used  in
\cite{pettre20032} in a computer animation context. Variants of this method
include  dynamic path reshaping  \cite{yoshida-humanoids05}: if  collisions appear
when animating  the locomotion  trajectory, it is  locally reshaped
and re-animated.  This two-stage  strategy does not guarantee that the
locomotion trajectory can be followed or that the local reshaping will
converge.

@@

Some planning methods for free-climbing robots \cite{bretl2006motion}
can be seen as a general way 
to consider quasi-static multi-step planning. They are not however directly applicable to
humanoid dynamically balanced locomotion.
Other recent contributions to the field of locomotion planning  include algorithms 
considering the dynamics at the planning phase \cite{shkolnik2011bounding}. This leads to 
a growth of algorithmic complexity, particularly costly for high-dimensional
systems such as humanoid robots. To the authors knowledge such techniques have not been
used on humanoid robotic platforms so far.
In our work, we show that it is not always
necessary to take into account dynamics while planning, even though the final
output of our algorithm is a dynamically balanced walk motion. This result was first
presented in \cite{dalibard2011small}.

@@

The main contribution of this work is a  whole-body motion planner for humanoid robots
that computes collision-free walking trajectories, based on
exact models of both robot and environment. It is used to solve manipulation tasks
that may require walking. The first stage of our algorithm uses a
constrained sampling-based motion 
planner and  computes a collision-free  statically balanced path  for a robot which can be fixed or sliding 
on the ground. Another  contribution of this  paper is the formal  proof that
this first path can always be  approximated by a dynamically balanced, collision-free
walking  trajectory. We have implemented this well-grounded 
method, and the results have been validated on the HRP-2 robot.

\section{Randomized Motion Planning on Constraint Manifolds}
\label{sec:wb}

FIXME: Avoid the mixup due to the use of task and constraints?

This section presents an algorithm for constrained motion planning. 
The problem solved here
differs from classic approaches in two ways:
\begin{enumerate}
\item the set of valid configurations is defined implicitly, as the ones satisfying a given 
  set of inverse kinematics and balance tasks;
\item the goal is also defined implicitly, by additional inverse kinematics tasks.
\end{enumerate}
During global planning, we will consider several types of
tasks and constraints for various reasons:
\begin{itemize}
\item Static balance: the CoM of the robot stays above
  the support polygon center, the two feet have fixed position and
  orientation.
\item End-effector position and orientation: the goals of some problems presented
  in the experimental section of this paper are defined as a specific robot hand pose,
  or a gaze direction.
\item Configuration task: our adaptation of randomized motion planning algorithms 
  uses tasks defined as the distance towards a given configuration in $\mathcal{C}$. 
  This will be detailed in the following section.
\end{itemize}

@@

The classic RRT algorithm, as presented in  \cite{kuffner00rrtconnect}, grows 
a random tree inside the robot 
collision-free configuration space 
$\mathcal{C}_{free}$. Each iteration of the algorithm attempts to extend the tree
by adding new vertices in the direction of a randomly selected configuration
$q_{rand}$. Algorithm~\ref{algRRT} shows the pseudo-code of the RRT algorithm.
It takes as input an initial configuration $q_0$ and grows a tree  $\mathcal{T}$ rooted 
in $q_0$.

@@

Section~\ref{sec:goal-sampling}  describes the goal manifold sampling, and 
Section~\ref{sec:extension} the adaptation or RRT random extensions to constrained motion planning.

@@

The way we generate a goal configuration is the following:
\begin{enumerate}
\item Shoot a random configuration in $\mathcal{C}$ with uniform distribution.
\item Call the \textit{constraint solver} (Algorithm~\ref{algo:newton}) on this configuration, with the static
  balance constraints and the goal task.
\item If success, check for collisions.
\end{enumerate}

@@

Fig.~\ref{fig:goal} shows different random configurations enforcing banlance 
constraints and solving a reaching task for an animated character.

@@

\caption{Random goal configurations solving a reaching task. All the
  configurations are balanced and collision-free, and the right hand of the character
  reaches the orange ball. Note that the ball is also an obstacle.}
\label{fig:goal}
\end{figure}

@@

We present in Fig.~\ref{fig:wb-shelves} an illustration of 
the use of randomized motion planning 
on complex manipulation problems. The humanoid robot HRP-2 faces shelves. It has to: (i) grasp
a ball lying on a shelf, (ii) put it on a higher shelf, (iii) come back to a natural
rest configuration. The statical balance manifold on which the planning takes place is
defined by tasks constraining the robot feet and CoM. The goal of the first problem

FIXME: what follows is not clear, rephrase to say that the hand
position is fixed and that it can rotate along the vertial axis.

is defined by a hand configuration task (3 positions and 2 orientations), and a gaze task
(the robot has to look at the ball). The goal of the second problem is defined by a 
different hand configuration and the gaze task.
Finally, the goal of the third problem is the rest configuration.

@@

\caption{HRP-2 grabs a ball on a shelf, puts it on another shelf, and comes back to 
  a rest position. Static balance constraints are enforced along the path, and 
  the intermediary goals consisting in grasping and displacing the ball are defined
  implicitly as inverse kinematics tasks.}
\label{fig:wb-shelves}
\end{figure}

We have run this set of motion planning problems 20 times. The results are compiled in Table~\ref{table:reaching}.
\begin{table}
\begin{tabular}{lrrrr}
& min & max & average & average \\ 
&&&& per problem \\
number of nodes & 43.00 & 481.00 & 102.70 \\
goal generation time \hspace {1.cm} & 1.00 & 1.56 & 1.22\\
planning time& 67.36 & 376.84 & 134.28 & 44.76\\
\end{tabular}
\caption {Experimental results on 20 runs: 3 motion planning problems and 2 goal generations. Time is expressed in seconds.}
\label{table:reaching}
\end{table}

@@

The previous section has presented a simple algorithm that solves manipulation planning
problems on a given constraint manifold of $\mathcal{C}$.

If we use this algorithm with static 
balance constraints without fixing globally the robot foot positions, it generates 
statically balanced paths for a robot sliding on the ground. Fig~\ref{fig:sliding} shows
an example of a whole-body collision-free path for a robot passing between two chairs.
Such paths are physically unfeasible. They are, however, easier to generate than feasible
dynamic trajectories because only geometric constraints are considered at planning time.

@@

  \caption{Collision-free  statically balanced path  for a  humanoid  robot  sliding on  the
    ground.}
  \label{fig:sliding}
\end{figure}

This section presents a \textit{constructive} proof that any such statically balanced, 
collision-free path for a legged robot sliding
on the ground can be approximated by a  dynamically balanced, collision-free walk trajectory.
The proof is based on ideas from control theory, in particular small-space 
controllability. It also uses the fact that balance criteria for dynamic walking are different
from the ones for static balance. An important point is that our result and method do not hold
for a biped robot that walks in a quasi-static way.

Section~\ref{sec:ssc} recalls the definition of small-space controllability and its
use in motion planning. Section~\ref{sec:humanoid-ssc} proves that a
dynamically walking legged robot is small-space controllable, while a quasi-statically
walking legged robot is not. Section~\ref{sec:ssc-application} shows how this property is used to
approximate collision-free statically balanced paths by dynamic walking trajectories.

@@

  \caption{The small-space controllability local property:  any configuration $q'$ 
    at a distance less than
    $\eta$ is reachable from $q$ by an admissible trajectory included in
    a ball of size $\epsilon$.}
  \label{fig:ssc1}
\end{figure}

The main  consequence of  this property  
in  motion planning  is the following theorem, that shows how planning for
dynamic systems is reduced to geometric planning
thanks to the small-space controllability property:

@@

Fig. \ref{fig:ssc2} shows an example of collision-free 
path approximation by admissible collision-free sub-trajectories. The convergence 
of this algorithm is guaranteed by the small-space
controllability property.

@@

\begin{proof}[Proof of Theorem \ref{thm:ssc}]
  Let $\mathcal{C}$ be the configuration space of a small-space controllable robot, and 
  $\mathcal{C}_{free} \subset \mathcal{C}$ the set of collision-free configurations. We
  consider in-contact configurations as colliding, so $\mathcal{C}_{free}$ is an open set.
  Let $\tau : [0,1] \rightarrow \mathcal{C}_{free}$ be a collision-free path. Thus for all $x \in [0,1]$,
  $\tau(x) \in \mathcal{C}_{free}$, there exists $\epsilon_x$ such that the open ball 
  $B(\tau(x),\epsilon_x)$ of center $\tau(x)$ and radius $\epsilon_x$ is included in 
  $\mathcal{C}_{free}$. The small-space controllability property states that for all $x$,
  there exists $\eta_x > 0$ such that every configuration $q \in B(\tau(x),\eta_x)$ is reachable 
  from $\tau(x)$ by a trajectory included in $B(\tau(x),\epsilon_x)$.

@@

This section  discusses a walking robot small-space controllability. To clarify
the presentation, we consider a simplified model of a legged robot consisting of two feet
of zero mass and a point mass free to move in three dimensions.
We do not consider 
the kinematic chains between the feet and the mass. The robot is walking on a flat terrain,
and the feet are assumed to have a positive surface. For our presentation, it is not 
necessary to consider the foot height, so the configuration space of the robot is:
\[
\mathcal{C} = SE(2) \times SE(2) \times \mathbb{R}^3
\]
It is of dimension 9.

@@

\begin{theorem}
\label{thm:humanoid-ssc}
A quasi-statically walking robot is not small-space controllable. A dynamically walking robot is.
\end{theorem}

@@

The first claim is straightforward. Let the robot be in a configuration $q$ where the two
feet are separated by a positive distance. Let  $L>0$ be the positive horizontal distance
between the CoM and the left foot (if the CoM is over the left foot, we can consider similarly 
the right foot). For all $\epsilon < L$, any valid trajectory starting from $q$, included in
the ball of center $q$ and radius $\epsilon$, is such that the CoM is never over the left foot.
Given the quasi-static walking conditions, the right foot of the robot is fixed along 
the trajectory. Thus, the set of accessible configurations from $q$ by staying inside 
$B(q,\epsilon)$ does not form a neighborhood of $q$, since it does not include any configuration
where the right foot has moved. This shows that the robot is not small-space controllable.

@@

Let us now consider a dynamically walking robot. Starting from any valid static configuration, 
the CoM can move vertically without affecting balance, so there is no need to 
consider this degree of freedom in the following. If the CoM is not over the edge of the support
polygon, it is possible to move it in a quasi-static way inside a neighborhood of its current position that projects itself over the support polygon.
It is thus sufficient and necessary to prove that for all $\epsilon >0$, it is possible to move
the feet while keeping the CoM inside a neighborhood of size $\epsilon$. Let such $\epsilon >0$
be arbitrarily fixed.

@@

Without loss of generality, let us assume that the robot is in a configuration
in which the CoM is at the horizontal position $(0,0)$, the foot centers  are 
aligned with the $y$-axis and the horizontal distance between the CoM and either of the foot centers
is $L$. To achieve dynamically balanced walking, we aim at making
$p_y(t)$ oscillate  between $-L$ and $L$. To move the ZMP  under a
given foot, only  the $y$ coordinate of the CoM  is of interest. Thus,
we will keep the $x$ coordinates  of the CoM and ZMP constant equal to
$0$. By hypothesis, the feet have a positive surface,
let $l>0$ be such that the length of the section of a foot along the 
$y$-axis is greater than $l$. Fig.~\ref{fig:simple-humanoid} summarizes the notations used
in the following.

@@

Starting from a static configuration at time $(t=0)$, we cannot apply
directly  a  command  $y(t)  =  \epsilon \sin(\omega  t)$  because  it
generates  a discontinuity  in the  speed of  the CoM  at time $(t=0)$. To
overcome this  discontinuity, we go through a  transient state between
$(t=0)$ and  $(t=T)$ for some  $T >0$. Let  $f:[0,T] \rightarrow
[0,1]$  be an  increasing function of class $C^\infty$  such  that  $f(0)  =  0$,
$\dot{f}(0) = 0$, $f(T) =  1$, $\dot{f}(T) = 0$ and $\ddot{f}(T)
=  0$.  We can explicitly  construct  such  an $f$   with  a
spline of degree  4.   We   also   request   that   for  all   $t   \in   [0,T]$,
$|2\epsilon\dot{f}(t)\frac{\omega}{\omega_0^2}|   \leq  \frac{l}{4}$
and   $|\epsilon\ddot{f}(t)/\omega_0^2|  \leq   \frac{l}{4}$.  These
inequalities  will be  used  to bound  the  trajectory of  the  ZMP. We  can
guarantee them by  choosing $T$ large enough. Let  us now consider the
following CoM motion:

@@

For    all     $t    \in    [0,T]$,    $f(t)     \epsilon    (1    +
\frac{\omega}{\omega_0}^2)  \sin  (\omega t)$  lies  between $-L$  and
$L$. The bounds on the derivatives of $f$ guarantee that $p_y(t)$ lies
between  $-L- l/2$ and  $L+ l/2$,  which means  that the  ZMP stays
inside  the  support  polygon.  Fig.  \ref{fig:zmp-inplace}  shows  an
example  of CoM  motion  on the  $y$  axis and  the corresponding  ZMP
motion. Once in permanent in-place walking state, the robot can come back
to  a static  state by  applying a  symmetric transient  state  used to
decrease  gradually  the amplitude  of  the  oscillations  of the  CoM
without  generating a  discontinuity in  the first  derivative  of the
command.

@@

\subsubsection{Remarks}
\paragraph{Generalization to a complete model:} More work is needed to generalize 
the previous proof to any legged robot model. First, because we use a model with
a point mass, we did not have to consider the angular momentum of the robot
around the CoM, while it accounts for dynamic balance. Moreover, when considering
an actual kinematic and dynamic model,
the positions of feet and CoM are not controlled DoFs, but inputs 
for inverse kinematic tasks. Continuity and regularity results about the robot
controller -- i.e. how much the configuration changes when the input of an inverse
kinematics task changes--, would be needed to generalize the proof. We have not
proven such properties for the inverse kinematics controller we use. 
The  proof presented above still gives the intuition for the
algorithm presented in the following.

@@

In the problem shown in Fig. \ref{fig:shelf} the robot has to put a ball on a
shelf, in a constrained apartment environment. The final configuration is defined 
implicitly as a desired hand position. We have generated automatically goal configurations 
solving the task, as described in Section~\ref{sec:wb} . Then, we have
applied our planner to generate a whole-body walking motion that solves the hand reaching
task.
